Higher-Order unification for free!

### 2024-04-09 paper/main.tex
delenda est

% \title{HO unification from object language to meta language}
\title{Higher order unification for free!}
\subtitle{Reusing the meta-language unification for the object language}

@@

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:\\
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
We defer the description of \elpiIn{beta} to \cref{sec:grounwork}, for now
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app[F|L]}>>.

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
\end{definition}

@@

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor put in $\sigma$ an application whose head is flexible
(an application node that could be erased by a $\beta$-reduction).
\todo{dire che W preservation is key for our claims}

% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) terms in
% \maybeeta, \maybebeta or \notllambda that were moved out of the way
% during compilation.

\section{Ground work for \Fo{}, \Ho and the compiler}
\label{sec:grounwork}
\todo{maybe move here napp beta deref fderef move}

We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

\input{code/pattern_fragment}

\noindent
Natural numbers represent the memory addresses that
identify unification variables in both languages.
The memory and its associated operations are described below:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\noteE[inline]{is new used?}

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it to enable the instantiation of a single
solution to different scopes. This is obtained via the \elpiIn{inctx}
container, and in particular via its \elpiIn{abs} binding constructor.
On the contrary a solution to a \Fo variable is a plain term.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO: explain.

@@

@@

\noindent
The compiler establishes a mapping between variables of the two languages.
In order to preserve invariant \ref{inv:uvaarity} we store the
arity of each \elpiIn{hvariable} in the mapping and we reuse an existing
mapping only if the arity matches.

\noteD{add ref to \cref{sec:invariant1}}

@@

\noindent
When a single \elpiIn{fvariable} occurs multiple times with different numbers
of arguments the compiler generates multiple mappings for it, on a first
approximation, and then makes the mapping bijective by introducing
\linketa; this detail is discussed in section \ref{sec:eta}.

As we mentioned in section~\ref{sec:nutshell} the compiler
replaces terms in \maybeeta and \notllambda with fresh variables
linked to the problematic terms. Each class of problematic terms
has a dedicated link.

@@

%\input{code/move}

\input{code/beta_fo}

\noindent
The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
variables).\footnote{Elpi provides it as a builtin, but one could implement it by
systematically loading the hypothetical rule \elpiIn{name x} every time
a nominal constant is postulated via \elpiInFN{pi x\ }}
The choice of using n-ary application, rather than binary, is to make it
easy to access the application's head. The price we pay is that substituting
an application in the head of an application should be amended by
``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
\footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
we do in this presentation and, as we explain later,
no equivalent of \elpiIn{napp} is needed in \Ho.}
Finally note that the cut operator is inessential, it could be
removed at the cost of a verbose test on the head of \elpiIn{L}
in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
can be \elpiIn{fcon}, \elpiIn{flam} or a name.

Applying the substitution corresponds to dereferencing a term with respect to
the memory.

To ease the comparison we split \Fo dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, and has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
has no corresponding operation in \Ho. The reasons for this asymmetry is
that an \elpiIn{fapp} node with a flexible head is always mapped
to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
preventing nested applications to materialize.

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}
that we shall define later.\todo{fixup}

>>>>

% \title{HO unification from object language to meta language}
\title{Higher-Order unification for free!}
\subtitle{Reusing the meta-language unification for the object language}

@@

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:\\
%
\vspace{4pt}
{
\setlength{\tabcolsep}{2pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
}
\vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app[F|L]}>>.

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
\end{definition}

@@

\noindent
\cref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
the problematic terms moved on the side by the compiler are not
put back by \hstep, hence \Ue{} can operate properly.
In \cref{sec:compilation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybeeta, \maybebeta and
\notllambda and how progress takes care of them preserving \wellb
and granting \cref{prop:simulation,prop:fidelity}.

Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
since decompilation can introduce (actually restore) problematic terms.

\section{Ground work for the compiler}
\label{sec:grounwork}

% The predicate to test this condition is called \elpiIn{pattern-fragment}:

% \input{code/pattern_fragment}

Unification variables are identified by a
natural number, that represents a memory addresses
The memory and its associated operations are described below:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.

Since each occurrence of a \Ho unification variables has a scope,
its solution needs to be abstracted over it to enable the
instantiation of a single assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.
On the contrary a solution to a \Fo variable is a plain term.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
The compiler establishes a mapping between variables of the two languages.

@@

Each \elpiIn{hvariable} is stored in the mapping together with
its arity so that the code of \ref{clause:malloc} below can preserve:

@@

@@

\noindent
When a single \elpiIn{fvariable} occurs multiple times with different numbers
of arguments the compiler generates multiple mappings for it, on a first
approximation, and then makes the mapping bijective by introducing
\linketa; this detail is discussed in section \ref{sec:invariant1}.

\noindent
As we mentioned in section~\ref{sec:nutshell} the compiler
replaces terms in \maybeeta, \maybebeta and \notllambda with fresh
variables linked to the problematic terms. Terms in \maybebeta do not
need a link since \Ho{} variables faithfully represent
the problematic term thanks to their scope.

@@

% \input{code/beta_fo}

% \noindent
% The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
% variables).\footnote{Elpi provides it as a builtin, but one could implement it by
% systematically loading the hypothetical rule \elpiIn{name x} every time
% a nominal constant is postulated via \elpiInFN{pi x\ }}
% The choice of using n-ary application, rather than binary, is to make it
% easy to access the application's head. The price we pay is that substituting
% an application in the head of an application should be amended by
% ``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
% \footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
% we do in this presentation and, as we explain later,
% no equivalent of \elpiIn{napp} is needed in \Ho.}
% Finally note that the cut operator is inessential, it could be
% removed at the cost of a verbose test on the head of \elpiIn{L}
% in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
% can be \elpiIn{fcon}, \elpiIn{flam} or a name.

Applying the substitution corresponds to dereferencing a term with respect to
the memory. To ease the comparison we split \Fo{} dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, and has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
has no corresponding operation in \Ho, and only ensures that
terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
<<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
and \elpiIn{L2}. The reasons for this asymmetry is
that an \elpiIn{fapp} node with a flexible head is always mapped
to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
preventing nested applications to materialize.

@@
